Continuous time

### 2023-09-08 continuous-time-draft.qmd
re slopes YOLO

```{r}
#| label: packages
#| cache: false
library(patchwork)
library(tidyverse)
library(lme4)
library(knitr)
library(emmeans)
```

@@

### Fit model
Reset duration within each session
```{r}
dat <- dat %>% 
  mutate(
    duration = duration - duration[1],
    hours = duration / 60,
    .by = c(pid, session)
  ) %>% 
  filter(prompt == "Wellbeing")
```

Fit linear change
```{r}
library(rms)
fit0 <- lmer(mood ~ hours + (1 | session) + (1 | pid), data = dat)
summary(fit0)
```

What's the distribution of session lengths
```{r}
median_ses_len <- dat %>% 
  group_by(pid, session) %>% 
  summarise(session_length = max(duration)) %>% 
  group_by(pid) %>% 
  summarise(median_len = median(session_length))

median(median_ses_len$median_len)

mean(median_ses_len$median_len > 60)

"Typical" session length is 24 minutes. Few participants typically play for more than an hour. Plot estimated trend over 0 to 2 hours.

@@

Fit model with only random intercepts.
```{r}
fit1 <- lmer(mood ~ rcs(hours, 5) + (1 | session) + (1 | pid), data = dat)
emm1 <- emmeans(
  fit1, 
  ~hours, 
  at = list(hours = seq(0,2, length.out = 100)),
  lmer.df = "asymptotic"
  )
p1 <- as.data.frame(emm1) %>% 
  ggplot(aes(hours, emmean)) + 
  geom_line() +
  geom_ribbon(aes(ymin = asymp.LCL, ymax = asymp.UCL), alpha = 0.25) +
  labs(x = "Session duration (hours)")
```

Adding `rcs(hours, 5)` as random effects would be messy, so let's fit the model with a linear random slope at the participant level. I don't think random slopes at the session level is needed? Moreover, I would expect that the exact random slopes specification have little impact on fixed effects in this case.
```{r}
path <- "models/lmm-fit2.Rds"
if (file.exists(path)) {
  fit2 <- read_rds(path)
} else {
  fit2 <- lmer(mood ~ rcs(hours, 5) + (1 | session) + (1 + hours | pid), data = dat)
  write_rds(fit2, path)
}
emm2 <- emmeans(
  fit2, 
  ~hours, 
  at = list(hours = seq(0,2, length.out = 100)),
  lmer.df = "asymptotic"
  )
p2 <- as.data.frame(emm2) %>% 
  ggplot(aes(hours, emmean)) + 
  geom_line() +
  geom_ribbon(aes(ymin = asymp.LCL, ymax = asymp.UCL), alpha = 0.25) +
  labs(x = "Session duration (hours)")
```

@@

```{r}
p0 / p1
```

@@

>>>>

```{r}
#| label: packages
#| cache: false
library(patchwork)
library(tidyverse)
library(lme4)
library(knitr)
library(emmeans)
library(splines)
```

@@

### Session lengths

What's the distribution of session lengths
```{r}
median_ses_len <- dat %>% 
  group_by(pid, session) %>% 
  mutate(
    duration = duration - duration[1],
  ) %>% 
  summarise(session_length = max(duration)) %>% 
  group_by(pid) %>% 
  summarise(median_len = median(session_length))
median(median_ses_len$median_len)
mean(median_ses_len$median_len > 60)
```

"Typical" session length is 35 minutes. Few participants typically play for more than an hour. Plot estimated trend over 0 to 2 hours.

### Fit model
Reset duration within each session
```{r}
dat <- dat %>%
  mutate(
    ps = interaction(pid, session, sep = "_")
  )  %>% 
  filter(prompt == "Wellbeing") %>% 
  mutate(
    duration = duration - duration[1],
    hours = duration / 60,
    .by = c(pid, session)
  ) 
```

Fit linear change
```{r}
library(rms)
fit0 <- lmer(mood ~ hours + (1 | ps) + (1 | pid), data = dat)
```

@@

Fit model with only random intercepts.
```{r}
fit1 <- lmer(mood ~ rcs(hours, 5) + (1 | ps) + (1 | pid), data = dat)
emm1 <- emmeans(
  fit1, 
  ~hours, 
  at = list(hours = seq(0,2, length.out = 100)),
  lmer.df = "asymptotic"
  )

## for plotting knot placement
get_knot_coords <- function(fit) {
    knots <- attr(rcs(dat$hours, 5), "parms")
    knots_x <- rcs(knots, knots)
    b0 <- fixef(fit)[1]
    b <- fixef(fit)[-1]
    knots_y <- b0 + knots_x[,1] * b[1] + knots_x[,2] * b[2] + knots_x[,3] * b[3] + knots_x[,4] * b[4]
    data.frame(
        x = knots,
        y = c(knots_y)
    )
}
p1 <- as.data.frame(emm1) %>% 
  ggplot(aes(hours, emmean)) + 
  geom_line() +
  geom_ribbon(aes(ymin = asymp.LCL, ymax = asymp.UCL), alpha = 0.25) +
  geom_point(
    data = get_knot_coords(fit1),
    aes(x,y)
  ) +
  labs(x = "Session duration (hours)")
```

Adding `rcs(hours, 5)` as random effects would be messy, so let's fit the model with a linear random slope at the participant level. I don't think random slopes at the session level is needed? Moreover, I would expect that the exact random slopes specification have little impact on fixed effects in this case.
```{r}
path <- "models/lmm-fit2.Rds"
if (file.exists(path)) {
  fit2 <- read_rds(path)
} else {
  fit2 <- lmer(mood ~ rcs(hours, 5) + (1 | ps) + (1 + hours | pid), data = dat)
  write_rds(fit2, path)
}
emm2 <- emmeans(
  fit2, 
  ~hours, 
  at = list(hours = seq(0,2, length.out = 100)),
  lmer.df = "asymptotic"
  )
p2 <- as.data.frame(emm2) %>% 
  ggplot(aes(hours, emmean)) + 
  geom_line() +
  geom_ribbon(aes(ymin = asymp.LCL, ymax = asymp.UCL), alpha = 0.25) +
  # geom_point(
  #   data = get_knot_coords(fit2),
  #   aes(x,y)
  # ) +
  labs(x = "Session duration (hours)")
```

@@

```{r}
p0 / p1 / p2
```

@@

## YOLO model with random slopes
Here I add a random slope to each piece of the spline, i.e. 4 random slopes at the participant level. The model is fit on a random subset of 1000 pids.
```{r}
# include 1000 participants
set.seed(1337)
pids <- sample(unique(dat$pid), 1000)
tmp <- filter(dat, pid %in% pids)
# so predict won't complain
tmp <- tmp %>% 
    mutate(
        pid = as.integer(pid),
        ps = as.integer(ps)
        )
# same spline but different parameterization
# usually less estimation problems
spline <- ns(
    tmp$hours, 
    knots = c(0.12, 0.4, 1), 
    Boundary.knots = c(0.016, 3)
)
colnames(spline) <- paste0("hours", 0:3)
tmp <- cbind(tmp, spline)
path <- "models/lmm-fit3.Rds"
if (file.exists(path)) {
  fit3 <- read_rds(path)
} else {
  fit3 <- lmer(
    mood ~ hours0 + hours1 + hours2 + hours3 + (1 | ps) + (1 + hours0 + hours1 + hours2 + hours3 | pid), 
    data = tmp
  )
  write_rds(fit3, path)
}

I'm interested in if the spline changes based on the pre-session scores. Let's plot the fixed effects at -2, -1, 0 times the random intercept SD + the corresponding random slope values based on their correlation with the random intercept.

```{r}
get_trend <- function(fit, data, re_int = 0, re_s0 = 0, re_s1 = 0, re_s2 = 0, re_s3 = 0) {
    hours <- seq(0,2, length.out = 100)
    x <- ns(
        hours, 
        knots = c(0.12, 0.4, 1), 
        Boundary.knots = c(0.016, 3)
    )
    b0 <- fixef(fit)[1] + re_int
    b <- fixef(fit)[-1] 
    y <- b0 + x[,1] * (b[1] + re_s0) + x[,2] * (b[2] + re_s1) + x[,3] * (b[3] + re_s2) + x[,4] * (b[4] + re_s3)
    data.frame(
        x = hours,
        y = c(y),
        re_int
    )
}
re <- as.data.frame(VarCorr(fit3))
pid_int <- re[2, "sdcor"]
pid_slope0 <- re[3, "sdcor"]
pid_slope1 <- re[4, "sdcor"]
pid_slope2 <- re[5, "sdcor"]
pid_slope3 <- re[6, "sdcor"]
pid_cor0 <- re[7, "sdcor"]
pid_cor1 <- re[8, "sdcor"]
pid_cor2 <- re[9, "sdcor"]
pid_cor3 <- re[10, "sdcor"]
sd1 <- 1
sd2 <- 2
rbind(
    get_trend(fit3, data = tmp),
    get_trend(
        fit3, 
        data = tmp,
        -sd1 * pid_int,
        -sd1 * pid_cor0 * pid_slope0, 
        -sd1 * pid_cor1 * pid_slope1, 
        -sd1 * pid_cor2 * pid_slope2, 
        -sd1 * pid_cor3 * pid_slope3
    ),
    get_trend(
        fit3, 
        data = tmp,
        -sd2 * pid_int,
        -sd2 * pid_cor0 * pid_slope0, 
        -sd2 * pid_cor1 * pid_slope1, 
        -sd2 * pid_cor2 * pid_slope2, 
        -sd2 * pid_cor3 * pid_slope3
        )
    ) %>% 
  ggplot(
    aes(
        x, 
        y, 
        color = factor(re_int), 
        group = factor(re_int)
        )
    ) + 
  geom_line() +
  labs(x = "Session duration (hours)")
```

The random intercepts are only weakly correlated with each piece of the cubic spline, so the shape doesn't change much.

Let's also predict some individual slopes.

```{r}
pred <- filter(tmp, pid %in% sample(unique(tmp$pid), 10))
pred$yhat <- predict(fit3, newdata = pred)
ggplot(
    data = pred,
    aes(
        x = hours,
        y = yhat,
        group = ps
    )
) + 
    geom_line(
        color = "blue", 
        linewidth = 1
    ) +
    geom_point(
        aes(y = mood)
    ) +
    facet_wrap(~pid)
```

Zoom in on pid 9263.
```{r}
ggplot(
    data = filter(pred, pid == 9263),
    aes(
        x = hours,
        y = yhat,
        group = ps
    )
) + 
    geom_line(
        aes(y = mood),
        linetype = "dashed",
        alpha = 0.33
    ) +
    geom_line(
        color = "blue", 
        linewidth = 2
        ) +
    geom_point(
        aes(y = mood)
    ) +
    facet_wrap(~session)
```

Probably good enough, even tho the session effect is only modelled using a random intercept.

### Predict change
Another YOLO approach to estimate individual change during a session is to to look at the model predictions.

TODO: should be based on the actual session length, not only the range of WB prompts.

```{r}
pred <- fit3@frame
pred$yhat <- predict(fit3)
pred %>% 
  # compare the last and first value
  dplyr::summarize(change = last(yhat) - yhat[1], .by = c(pid, ps)) %>%
  # average over sessions within participants
  dplyr::summarize(change = mean(change), .by = pid) %>% 
  # average over participants
  dplyr::summarize(mean(change), sd(change))
```

We can also plot the distribution of the participants average change during sessions.
```{r}
pred %>% 
  dplyr::summarize(change = last(yhat) - yhat[1], .by = c(pid, ps)) %>%
  dplyr::summarize(change = mean(change), .by = pid) %>% 
  ggplot(aes(change)) + geom_histogram()
```

They really don't change much. Let's look at a participant out in the tails.
```{r}
pred %>% 
  dplyr::summarize(change = last(yhat) - yhat[1], .by = c(pid, ps))  %>% 
  filter(change > 0.1) %>% head
```

Zoom in on pid 23.
```{r}
pred2 <- filter(tmp, pid == 23)
pred2$yhat <- predict(fit3, newdata = pred2)
ggplot(
    data = pred2,
    aes(
        x = hours,
        y = yhat,
        group = ps
    )
) + 
    geom_line(
        aes(y = mood),
        linetype = "dashed",
        alpha = 0.33
    ) +
    geom_line(
        color = "blue", 
        linewidth = 2
        ) +
    geom_point(
        aes(y = mood)
    ) +
    facet_wrap(~session)
```

This look kinda reasonable?
